---
output: github_document
---

# Tidyup 2: tidyr \<-> stringr alignment

**Champion**: Hadley Wickham  
**Co-champion**: Mine Çetinkaya-Rundel  
**Status**: Draft

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Abstract

tidyr provides a handful functions like `separate()` and `extract()` that perform string manipulation.
Unfortunately, these functions are only weakly connected to their stringr equivalents, making them harder to learn and remember.
This tidyup proposes a new cohesive family of string manipulation functions to tackle common problems that arise during data tidying.
The existing functions will remain so existing code is unaffected, but will be superseded to steer folks towards a more consistent API.

```{r setup}
library(tidyr)
library(stringr)
```

## Motivation

stringr and tidyr provide surprisingly different interfaces for string manipulation.
This means that common tasks require integrating two distinct syntaxes and learning more about stringr doesn't improve your use of tidyr.

The root cause of the problem is that stringr was designed before I understood the importance of data frames, so stringr functions that need to return multiple results use matrices rather than data frames.
This is particularly painful since dplyr 1.0.0 gave `summarise()` and `mutate()` the ability to create multiple columns from data frame results.

To overcome this weakness, tidyr provides a few functions that work directly with data frames.
These take a data frame and column specification, and return a modified data frame, typically with new columns specified by the call.
Unfortunately, the tidyr functions have little relationship to their stringr equivalents:

-   `extract()` is similar to `str_match()`.
-   `separate()` with regex is similar to `str_split_fixed()`.
-   `separate()` with numeric `col` (which splits by position) has no equivalent in stringr.
-   `separate_rows()` is equivalent to `str_split()`

Additionally, tidyr uses the base R regexp engine and syntax, not the stringr engine and syntax.
There are relatively few differences between the engines, but the syntax for controlling regular expression options (e.g. ignoring case) is very different.
In the past, we have kept stringr out of the dependencies of tidyr because it is a heavy dependency (because it can be expensive/painful to build stringi on linux).
However, RSPM makes stringi easier to install, so it would be nice to use it if available, following the footsteps laid by [Radix Ordering in](https://github.com/tidyverse/tidyups/blob/main/003-dplyr-radix-ordering.md) [`dplyr::arrange()`](https://github.com/tidyverse/tidyups/blob/main/003-dplyr-radix-ordering.md).

Additionally there are a few gaps in the current API:

-   Neither stringr nor tidyr provide a particularly nice way to work with named capture groups.
    There's also scope for more directly matching column names to patterns if `extract()` didn't attempt to match the syntax of `separate()` so closely, e.g. `str_extract(x, c(type = ".*", "|", fruit = "orange|apple"))`.

-   tidyr only provides one function for splitting a string into rows, not columns: `separate_rows()`.
    What other tidyr functions should create rows?

-   `tidyr::separate()` can split up a string by position or by pattern.
    This now seems too clever, and makes it hard to notice that there are two different tools here.

## Solutions

Finding a solution to this problem first requires a careful analysis of the cases.
I think we can break splitting down into two largely independent components:

-   How do you specify the pieces?

    -   By length/position, e.g. `c(1, 5, 10)`.
    -   By a regular expression pattern, e.g. `","`, `"\s+"`.
    -   With regular expression "groups" formed by `()`,`"a(b)c".`

-   How do the piece vary from element to element?

    -   Does each string have the same number of components which we might name and put in columns?

    -   Does each string potential have a varying number of components which we'd put into rows?

    (We don't need to consider splitting into rows and columns simultaneously because a compound problem can always be solved by first splitting into rows and then into columns.)

Since there is no way to split up by regexp group and produce a variable number of pieces, these two components define five ways to split up a string.
The following table matches them up with existing stringr and tidyr functions:

| By      | Into            | stringr         | tidyr             |
|---------|-----------------|-----------------|-------------------|
| Length  | Named/fixed     |                 | `separate()`      |
| Length  | Unnamed/varying |                 |                   |
| Pattern | Named/fixed     | `str_split_n()` | `separate()`      |
| Pattern | Unnamed/varying | `str_split()`   | `separate_rows()` |
| Group   | Named/fixed     | `str_match()`   | `extract()`       |

Note that `separate()` is shown in two locations, and there are a number of empty cells.
Interestingly, there's no stringr function that allows you to extract multiple pieces by position; you currently have to call `str_sub()` multiple times yourself.

So solving this challenge is primarily about replacing this inconsistent and incomplete table with a consistent and complete set of functions.

### Function names

I propose the following set of function names:

| By      | Into            | stringr                      | tidyr                    |
|---------|-----------------|------------------------------|--------------------------|
| Length  | Named/fixed     | `str_separate_at_wider()`    | `separate_at_wider()`    |
| Length  | Unnamed/varying | `str_separate_at_longer()`   | `separate_at_longer()`   |
| Pattern | Named/fixed     | `str_separate_by_wider()`    | `separate_by_wider()`    |
| Pattern | Unnamed/varying | `str_separate_by_longer()`   | `separate_by_longer()`   |
| Group   | Named/fixed     | `str_separate_group_wider()` | `separate_group_wider()` |

These names are guided by the following thinking:

-   There should be a simple and consistent transformation between the stringr and tidyr pairs.
    stringr functions reliably start with `str_`.
    This suggests the tidyr functions should be the same as the stringr function with the `str_` stripped off.

-   Suffixes should distinguish between the different ways of splitting.
    `at` and `by` work well here since there are two main cases (position and pattern) but would start to get vague if we have too many other categories.
    `_at` family was deprecated in dplyr so some possibility for confusion; but it's probably fairly clear that the context is different here.

    I don't love `group`, because I don't think many people think about regexp `()` as creating groups (even though that's the technical name), but there doesn't seem to be any obvious better word.
    This function also feels a little different to "splitting" which is what lead to `tidyr::extract()` but I think it's better to keep the root verb the same since there are otherwise so many similarities.

-   A second suffix should distinguish between splitting into fixed/named and varying/unnamed.
    If we assume a strong connection between fixed/named and columns and varying/unnamed and rows, the obvious suffixes are `wider` and `longer`.

    The return type associated with these suffixes will be a little different between tidyr and stringr.
    For stringr, `wider` implies a data frame and `longer` implies a list of character vectors.
    For tidyr, `wider` implies growing the columns and `longer` implies growing the rows.

    This will also influence the function arguments since `wider` functions names.

-   Altogether this implies we need one verb to represent this whole family.
    `separate()` seems like the natural choice since it is it the closest existing function and should be familiar to many tidyverse users.
    I initially considered `split` since it's shorter and has a strong connection to `str_split()`, but `base::split()` means something fairly different.
    There's something nice about an s verb for tidyr since it feels "string-y".

    For the purpose of completeness, other unused split/separate synonyms: fracture, cleave, snap, splinter, divide, sever, rend, dissect.

### Eliminate tidyr functions

Before we continue on to other design considerations, it's worth double checking that we can't eliminate the tidyr function in favour of using dplyr and stringr directly.
This seemed like the "obvious" solution to me at least twice, so I wanted to make sure why it doesn't work.

To make the discussion concrete I'll start with quick and dirty implementations of `str_separate_by_wider()` and `str_separate_by_longer()`:

```{r}
str_separate_by_wider <- function(x, pattern, into) {
  mat <- str_split_fixed(x, pattern, length(into))
  colnames(mat) <- into
  tibble::as_tibble(mat)
}
str_separate_by_longer <- function(x, pattern) {
  str_split(x, pattern)
}

str_separate_by_wider(c("a-b", "c-d"), "-", c("first", "second"))
str_separate_by_longer(c("a-b", "c-d-e"), "-")
```

We can use `str_separate_by_wider()` directly with `mutate()` to achieve the same affect as `separate()`:

```{r}
library(dplyr, warn.conflicts = FALSE)

df <- tibble(x = c("a-b", "c-d"))
df %>% separate(x, c("first", "second"), "-")

df %>% 
  mutate(
    str_separate_by_wider(x, "-", c("first", "second")), 
    .keep = "unused"
  )
```

This works because `mutate()` will create multiple columns if an unnamed calculation returns a data frame.

However, there's no straightforward way to use `str_separate_by_longer()` to create multiple columns.
The simplest pure dplyr approach is probably to use `rowwise():`

```{r}
df %>%
  rowwise() %>% 
  summarise(str_separate_by_longer(x, "-")[[1]])
```

But it's quite hard to explain why this works, and the use of `summarise()` means that you need to explicitly list any variables that you want to leave as is.
This possibly speaks to a missing verb in dplyr (e.g. <https://github.com/tidyverse/dplyr/issues/5874>) but it's also possible that creating rows is fundamentally a tidyr operation.
Either way, there's no easy fix in sight, and it seems unappealing to force folks to learn sophisticated dplyr idioms in order to tackle a routine data tidying task.

### Arguments

-   `str_separate_*_*()`: first argument is character vector.

-   `str_*_*_wider()` needs a `col_names` argument and arguments that controls what happens if there are too many/too few pieces.
    Probably needs `names_sep` for consistency with `unnest_wider()`.

```{=html}
<!-- -->
```
-   `str_separate_group_wider()` has a few options for specifying names:

    -   Named groups `"(?<type>.*)|(?<fruit>orange|apple)`

    -   Named vector: `c(type = ".*", "|", fruit = "orange|apple"))`

-   Should `convert` continue to exist?
    Should it use `type.convert()` or `readr::type_convert()`?
    How to override imputed types?

The vast majority of the work will occur in the stringr functions, and the tidyr functions will inherit their arguments.
`separate_*_*()` will work similarly by instead of taking character vector, takes pair of data frame + tidyselect specification.
Also has `remove` argument which controls whether or not the input column is removed.
The tidyr functions would primarily be wrappers that take care of removing the old column and adding the new rows/columns to the input data frame.
tidyr would require a new helper similar to [`inflate()`](https://github.com/tidyverse/dplyr/pull/5888).
`tidyr::uncount()` and `tidyr::separate_rows()` would be rewritten to use this new function.

### Conditional stringr dependency

Use stringr regexps if stringr is installed, or pattern wrapped with `stringr::regexp()` or friends.
If stringr not installed, message that using base R; provide `perl()` or similar to suppress that message if you deliberately want to use PCRE.

Or we could just say that the new functions require stringr + stringi, and see what the fallout is like.

But where would the implementations of base R regexp live?
A bit unappealing to put in tidyr.
Use stringrb?
Particularly code for handling too many/too few matches.

## Implications for teaching

Need to update R4DS.
