---
output: github_document
---

# Tidyup 2: tidyr \<-> stringr alignment

**Champion**: Hadley Wickham  
**Co-champion**: Mine Çetinkaya-Rundel  
**Status**: Draft

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Abstract

tidyr provides a handful of functions like `separate()` and `extract()` that perform string manipulation.
Unfortunately, these functions are only weakly connected to their stringr equivalents, making them harder to learn and remember.
This tidyup proposes a new cohesive family of string manipulation functions to tackle common problems that arise during data tidying.
The existing functions will remain so existing code is unaffected, but will be superseded to steer folks towards a more consistent API.

```{r setup}
library(tidyr)
library(stringr)
```

## Motivation

stringr and tidyr provide surprisingly different interfaces for string manipulation.
This means that common tasks require integrating two distinct syntaxes and learning more about stringr doesn't help your use of tidyr.

The root cause of the problem is that stringr was designed before I understood the importance of data frames, so stringr functions that need to return multiple values use matrices rather than data frames.
This is particularly unfortunate given that `dplyr::summarse()` and `mutate()` can now create multiple columns from data frame results.

To overcome this weakness, tidyr provides family of string manipulation functions that work with data frames.
Unfortunately, the names of the tidyr functions have little relationship to their stringr equivalents:

-   `extract()` creates new variables from groups, and is similar to `str_match()`.
-   `separate(sep = "pattern")` creates new variables by splitting with a regular expression, and is similar to `str_split_fixed()`.
-   `separate(sep = c(2, 5, 10)`) creates new variables by splitting based on position, and has no equivalent in stringr.
-   `separate_rows()` creates new rows, and is similar to `str_split()`.

Additionally, tidyr uses the base R regexp engine and syntax, not the stringr engine and syntax.
There are relatively few differences between the engines, but the syntax for controlling regular expression options (e.g. ignoring case) is very different.
In the past, we have kept stringr out of tidyr dependencies because it can be expensive/painful to build stringi on linux.
However, most tidyverse users will have stringr installed, and RSPM makes stringi easier to install, so it would be nice to use it if available, following the footsteps laid by [Radix Ordering in](https://github.com/tidyverse/tidyups/blob/main/003-dplyr-radix-ordering.md) [`dplyr::arrange()`](https://github.com/tidyverse/tidyups/blob/main/003-dplyr-radix-ordering.md).

Additionally there are a few gaps in the current API:

-   Neither stringr nor tidyr provide a particularly nice way to work with named capture groups.
    There's also scope for more directly matching column names to patterns if `extract()` didn't attempt to match the syntax of `separate()` so closely, e.g. `str_extract(x, c(type = ".*", "|", fruit = "orange|apple"))`.

-   tidyr only provides one function for splitting a string into rows, not columns: `separate_rows()`.
    What other tidyr functions should create rows?

-   `separate()` can split up a string by position or by pattern.
    This now seems too clever, and makes it hard to notice that there are two different tools here.

## Solution

Finding a solution to this problem first requires a careful analysis of the possible inputs.
I think we can break splitting down into two largely independent components:

-   How do you specify the pieces? There are three common cases:

    -   By width or position e.g. `c(1, 3, 2)`/`c(1, 4, 6)`. This typically comes from fixed width formats.
    -   By a regular expression pattern, e.g. `","`, `"\s+"`. This often comes from `str_flatten(x, "|")` or similar.
    -   With regular expression "groups" formed by `()`,`"a(b)c"`. This handles the remaining cases.

-   How do the pieces vary from element to element?

    -   Does each string have the same number of components which we might name and put in columns?

    -   Does each string potentially have a varying number of components which we'd put into rows?

    (We don't need to consider splitting into rows and columns simultaneously because a compound problem can always be solved by first splitting into rows and then into columns.)

In principle, this should define 6 (3 \* 2) situations, but regexp groups always produce a fixed number of matches so one case does not exist.
The following table matches the five cases up with existing stringr and tidyr functions:

| By      | Into            | stringr         | tidyr             |
|---------|-----------------|-----------------|-------------------|
| Length  | Named/fixed     |                 | `separate()`      |
| Length  | Unnamed/varying |                 |                   |
| Pattern | Named/fixed     | `str_split_n()` | `separate()`      |
| Pattern | Unnamed/varying | `str_split()`   | `separate_rows()` |
| Group   | Named/fixed     | `str_match()`   | `extract()`       |

Note that `separate()` is shown in two locations, and there are a number of empty cells.
Interestingly, there's no stringr or tidyr function that allows you to extract multiple pieces by position; you currently have to call `str_sub()` multiple times yourself.

So solving this challenge is primarily about replacing this inconsistent and incomplete table with a consistent and complete set of functions.

### One function per case

I propose the following set of function names:

| By      | Into            | stringr                      | tidyr                    |
|---------|-----------------|------------------------------|--------------------------|
| Length  | Named/fixed     | `str_separate_at_wider()`    | `separate_at_wider()`    |
| Length  | Unnamed/varying | `str_separate_at_longer()`   | `separate_at_longer()`   |
| Pattern | Named/fixed     | `str_separate_by_wider()`    | `separate_by_wider()`    |
| Pattern | Unnamed/varying | `str_separate_by_longer()`   | `separate_by_longer()`   |
| Group   | Named/fixed     | `str_separate_group_wider()` | `separate_group_wider()` |

These names are guided by the following thinking:

-   There should be a simple and consistent transformation between the stringr and tidyr pairs.
    stringr functions reliably start with `str_`.
    This suggests the tidyr functions should be the same as the stringr function without the `str_` prefix.

-   Suffixes should distinguish between the different ways of splitting.
    `at` and `by` work well here since there are two main cases (position and pattern) but would start to get vague if we have too many other categories.
    The `_at` family was deprecated in dplyr so some possibility for confusion; but it's probably fairly clear that the context is different here.

    I don't love `group`, because I don't think many people think about regexp `()` as creating groups (even though that's the technical name), but there doesn't seem to be any obvious better word.
    This function also feels a little different to "splitting" which is what lead to `tidyr::extract()` but I think it's better to keep the root verb the same since there are otherwise so many similarities.

-   A second suffix should distinguish between splitting into fixed/named and varying/unnamed.
    If we assume a strong connection between fixed/named and columns and varying/unnamed and rows, the obvious suffixes are `wider` and `longer`.

    The return type associated with these suffixes will be a little different between tidyr and stringr.
    For stringr, `wider` implies a data frame and `longer` implies a list of character vectors.
    For tidyr, `wider` implies growing the columns and `longer` implies growing the rows.

    This will also influence the function arguments since `wider` functions also need column names.

-   Altogether this implies we need one verb to represent this whole family.
    `separate()` seems like the natural choice since it is it the closest existing function and should be familiar to many tidyverse users.
    I initially considered `split` since it's shorter and has a strong connection to `str_split()`, but `base::split()` means something fairly different.
    There's something nice about an s verb for tidyr since it feels "string-y".

    For the purpose of completeness, other unused split/separate synonyms: fracture, cleave, snap, splinter, divide, sever, rend, dissect.

### Implementation

The vast majority of the work will occur in the stringr functions, with the tidyr functions being responsible for adding and removing (if `remove` is `TRUE`) columns.
`separate_*_longer()` will need a new helper similar to [`inflate()`](https://github.com/tidyverse/dplyr/pull/5888), which can also be used in `tidyr::uncount()` and `tidyr::separate_rows()`.

Other considerations:

-   `*_wider()` needs a `col_names` argument and arguments that controls what happens if there are too many/too few pieces.
    Probably needs `names_sep` for consistency with `unnest_wider()`.

```{=html}
<!-- -->
```
-   `str_separate_group_wider()` needs to be able to specify names inside the pattern:

    -   Named groups `"(?<type>.*)|(?<fruit>orange|apple)`

    -   Named vector: `c(type = ".*", "|", fruit = "orange|apple"))`

-   For `_wider()`, should `convert` continue to exist?
    Should it use `type.convert()` or `readr::type_convert()`?
    Can you override imputed types or do you need to switch to a different function?
    Or use `transform` argument like `unnest_wider()`?

### stringr dependency

Should these functions use stringr or not?
It's nice to keep dependencies low, but this will considerably increase the cost of implementation because they'll need both stringr and base R implementations.

Before committing to implementation a pure base R solution, I think we should discover if this is a real need.
To do this, I suggest we take stringr as a suggested dependency that's required by `separate_*()`.
We'll open an issue to track how much pain this causes for people, and consider a base R implementation in a future release if there's significant need.

## Alternatives

These alternatives were considered and rejected.

### One function per column

Another option is to divide the problem into two parts:

| Into            | stringr                 | tidyr               |
|-----------------|-------------------------|---------------------|
| Named/fixed     | `str_separate_wider()`  | `separate_wider()`  |
| Unnamed/varying | `str_separate_longer()` | `separate_longer()` |

Then each function has a `split` argument which can be one of the following cases:

-   `by_position(1, 2, 3)`
-   `by_width(1, 2, 3)`
-   `by_separator(",", ignore.case = TRUE)`
-   `by_regex_group("a(.)c")` (`by_group()` is too short/ambiguous).

(This could also be defined by 3-4 mutual exclusive arguments, but that is hard to extend and hard to document)

Pros:

-   8 functions instead of 10, and extending with more splitting types is easy.
-   Composing functions in this way seems more "tidyverse" like.

Cons:

-   Composing the functions on your own makes it harder to see the full set of options, and there's not enough total options to make the flexibility really worth it.
-   Will be tempting to use `someotherfunction(x, by_position(1, 2, 3))`.

Overall, this proposal only reduces the total number of functions by two, in return for a small gain.

### Eliminate tidyr functions

Can we eliminate the tidyr wrappers altogether and just use dplyr + stringr?
This seemed like the "obvious" solution to me at least twice, so I wanted to make sure to explain why it doesn't work.

To make the discussion concrete I'll start with quick and dirty implementations of `str_separate_by_wider()` and `str_separate_by_longer()`:

```{r}
str_separate_by_wider <- function(x, pattern, into) {
  mat <- str_split_fixed(x, pattern, length(into))
  colnames(mat) <- into
  tibble::as_tibble(mat)
}
str_separate_by_longer <- function(x, pattern) {
  str_split(x, pattern)
}

str_separate_by_wider(c("a-b", "c-d"), "-", c("first", "second"))
str_separate_by_longer(c("a-b", "c-d-e"), "-")
```

We can use `str_separate_by_wider()` directly with `mutate()` to achieve the same affect as `separate()`:

```{r}
library(dplyr, warn.conflicts = FALSE)

df <- tibble(x = c("a-b", "c-d"))
df %>% separate(x, c("first", "second"), "-")

df %>% 
  mutate(
    str_separate_by_wider(x, "-", c("first", "second")), 
    .keep = "unused"
  )
```

This works because `mutate()` will create multiple columns if an unnamed calculation returns a data frame.

However, there's no straightforward way to use `str_separate_by_longer()` to create multiple columns.
The simplest pure dplyr approach is probably to use `rowwise():`

```{r}
df %>%
  rowwise() %>% 
  summarise(str_separate_by_longer(x, "-")[[1]])
```

But it's quite hard to explain why this works, and the use of `summarise()` means that you need to explicitly list any variables that you want to leave as is.
This possibly speaks to a [missing verb](https://github.com/tidyverse/dplyr/issues/5874) in dplyr but it's also possible that creating rows is fundamentally a tidyr operation.
Either way, there's no easy fix in sight, and it seems unappealing to force folks to learn sophisticated dplyr idioms in order to tackle a routine data tidying task.

## Implications for teaching

Need to update R4DS.

## Questions
