---
output: github_document
editor_options: 
  chunk_output_type: console
---

# Tidyup 3: Radix Ordering in `dplyr::arrange()`

**Champion**: Davis

**Status**: Proposal

## Abstract

As of dplyr 1.0.6, `arrange()` uses a modified version of `base::order()` to sort columns of a data frame. Recently, vctrs has gained `vec_order()`, a fast radix ordering function with first class support for data frames and custom vctrs types, along with enhanced character ordering. The purpose of this tidyup is to propose switching `arrange()` from `order()` to `vec_order()` in the most user-friendly way possible.

## Motivation

Thanks to the data.table team, R \>= 3.3 gained support for an extremely fast radix ordering algorithm in `order()`. This has become the default algorithm for ordering most atomic types, with the notable exception of character vectors. Radix ordering character vectors is only possible in the C locale, but the shell sort currently in use by `order()` respects the system locale. Because R is justifiably hesitant to break backwards compatibility, if *any* character vector is present, the entire ordering procedure falls back to a much slower shell sort. Because dplyr uses `order()` internally, the performance of `arrange()` is negatively affected by this fallback.

Inspired by the performance of the radix ordering algorithm, and by its many practical applications for data science, a radix order-based `vec_order()` was added to vctrs, which has the following benefits:

-   Radix ordering on all atomic types.

-   First class support for data frames, including specifying sorting direction on a per column basis.

-   Support for an optional character transformation function, which generates an intermediate *sort key* that is unique to a particular locale. When sorted in the C locale, the sort key returns an ordering that would be equivalent to directly sorting the original vector in the locale that the sort key was generated for.

It is worth looking at a quick example that demonstrates just how fast this radix ordering algorithm is when compared against the defaults of `order()`.

```{r}
library(stringi)
library(vctrs) # r-lib/vctrs#1375
set.seed(123)

# 10,000 random strings, sampled to a total size of 1,000,000
n_unique <- 10000L
n_total <- 1000000L

dictionary <- stri_rand_strings(
  n = n_unique, 
  length = sample(1:30, n_unique, replace = TRUE)
)

x <- sample(dictionary, size = n_total, replace = TRUE)

head(x)

# Force `order()` to use the C locale to match `vec_order()`
bench::mark(
  base = withr::with_locale(c(LC_COLLATE = "C"), order(x)),
  vctrs = vec_order(x)
)

# Force `vec_order()` to use the American English locale, which is also
# my system locale. To do that we'll need to generate a sort key, which
# we can sort in the C locale, but the result will be like we sorted
# directly in the American English locale.
bench::mark(
  base = order(x),
  vctrs = vec_order(x, chr_transform = ~stri_sort_key(.x, locale = "en_US"))
)

# Generating the sort key takes most of the time
bench::mark(
  sort_key = stri_sort_key(x, locale = "en_US")
)
```

In dplyr, we'd like to utilize `vec_order()` while breaking as little code as possible. Switching to `vec_order()` has many potential positive impacts, including:

-   Improved performance when ordering character vectors.

    -   Which also results in improved overall performance when ordering character vectors alongside other atomic types, since it would no longer cause the whole procedure to fall back to a shell sort.

-   Improved reproducibility across sessions by ensuring that the default behavior doesn't depend on an environment variable, `LC_COLLATE`, which is used by `order()` to determine the locale.

-   Improved reproducibility across OSes where `LC_COLLATE` might differ, even for the same locale. For example, `"en_US"` on a Mac is approximately equivalent to `"English_United States.1252"` on Windows.

-   Improved consistency within the tidyverse. Specifically, with stringr, which defaults to the `"en"` locale and has an explicit argument for changing the locale, i.e. `stringr::str_order(locale = "en")`.

However, making this switch also has potential negative impacts:

-   Breaking code that relies strongly on the ordering resulting from using the `LC_COLLATE` environment variable.

-   Surprising users if the new ordering does not match the previous one.

## Proposal

To switch to `vec_order()` internally while surprising the least amount of users, it is proposed that the data frame method for `arrange()` gain a new argument, `.locale`, with the following properties:

-   Defaults to `dplyr_locale()`, see below.
-   If stringi is installed, allow a string locale identifier, such as `"fr"` for French, for explicitly adjusting the locale. If stringi is not installed, and the user has explicitly specified a locale identifier, an error will be thrown.
-   Allow `"C"` to be specified as a special case, which is an explicit way to request the C locale. If the exact details of the ordering are not critical, this is often much faster than specifying a locale identifier. This does not require stringi.

`dplyr_locale()` would be a new exported helper function. Its purpose is to return a string containing the default locale to sort with. It has the following properties:

-   If stringi is installed, the American English locale, `"en"`, is used as the default.

-   If stringi is not installed, the C locale, `"C"`, will be used. A warning will be thrown informing the user of this fallback behavior, and encouraging them to silence the warning by either installing stringi or explicitly specifying `.locale = "C"`.

-   Alternatively, to globally override the above default behavior, the global option, `tidyverse.locale_collation`, can be set to either `"C"` or a string locale identifier. Setting this to anything except `"C"` would require stringi.

American English has been chosen as the default solely because we believe it is the most used locale among R users, so it is the least likely to change existing results. That said, we understand that non-English speakers will not want to set the `.locale` argument *every time* they call `arrange()`, so the global option provides a way to change that default for their scripts. Feedback has indicated that while a global option can decrease reproducibility between sessions, in this case the benefits of it outweigh the costs.

On certain systems, stringi can be a difficult dependency to install. Because of this, this proposal recommends that stringi only be *suggested* so that users without stringi can still use dplyr.

This proposal relies on `stringi::stri_sort_key()`, which generates the sort key mentioned under Motivation as a proxy that can be ordered in the C locale. However, sort key generation is expensive. In fact, it is often the most expensive part of the entire sorting process. That said, generating a sort key + sorting it in the C locale is generally still 5-10x faster than using `order()` directly. If performance is critical, users can specify `.locale = "C"` to get the maximum benefits of radix ordering.

## Reference Implementation

-   Using `vec_order()` in `arrange()`, and adding `.locale`

    -   <https://github.com/tidyverse/dplyr/pull/5942>

-   Renaming `vec_order_radix()` to `vec_order()`

    -   <https://github.com/r-lib/vctrs/pull/1375>

## Backwards Compatibility

### arrange()

The proposal outlined above is purposefully as conservative as possible, preserving the results of programs using the American English locale, which is the most widely used locale in R, while sacrificing a bit of performance from the generation of the sort key.

That said, this proposal will impact non-English Latin script languages. For example, in a majority of Latin script languages, including `"en"`, ø sorts after o, but before p. However, a few languages, such as Danish, sort ø as a unique character after z. Danish users that have `LC_COLLATE` set to Danish may be surprised that `arrange()` would now be placing ø in the "wrong order" even though they have set that global option. The fix would be to either set `.locale = "da"` in their calls to `arrange()`, or to set `options(tidyverse.locale_collation = "da")` to override this default globally.

```{r, message=FALSE}
library(dplyr) # tidyverse/dplyr#5942
library(withr)

tbl <- tibble(x = c("ø", "o", "p", "z"))

# `"en"` default
arrange(tbl, x)

# Set the locale in `arrange()` directly
arrange(tbl, x, .locale = "da")
```

### arrange_at/if/all()

While these three variants of `arrange()` are superseded, we have decided to add a `.locale` argument to each of them anyways.

## Alternatives

### No global option

The original proposal defaulted to the American English locale, but did not include a way to globally override this. This had the benefit of being more reproducible across sessions, since as long as stringi was installed it was guaranteed that the locale was always American English unless specified otherwise through `.locale`. However, in the tidyverse meeting on 2021-06-14 it was determined that not including a way to globally override this default was too aggressive of a change for non-English Latin script users. With a large script containing multiple calls to `arrange()`, each call would have to be updated. Additionally, if a function from a package that a user didn't control called `arrange()` internally, then without a global option they would have no way to adjust the locale until that package author exposed the new `.locale` argument. Ultimately, in this case the benefits of the global option outweigh the potential reproducibility issues that might arise from it.

### Defaulting to the C locale

Another alternative is to default `arrange()` to the C locale, while still allowing users to specify `.locale` for ordering in alternative locales.

-   This has the benefit of making it clearer that stringi is an optional dependency, which would only be used if a user requests a specific locale identifier. The default behavior would never require stringi.

-   Additionally, the performance improvements would be even more substantial since no sort key would be generated by default.

-   However, this would have the potential to alter nearly every call to `arrange()`, since the C locale is not identical to the American English locale. In particular, in the C locale all capital letters are ordered before lowercase ones, such as: `c(A, B, a, b)` , while in the American English locale letters are first grouped together regardless of capitalization, and then lowercase letters are placed before uppercase ones, like: `c(a, A, b, B)`. This may look like a small difference, but it would surprise enough users to justify not defaulting to the C locale.

-   This is also not as consistent with stringr, which defaults to `"en"`.

### Tagged character vectors

A final proposed alternative is to implement a "tagged" character vector class, with an additional attribute specifying the locale to be used when ordering. This would remove the need for any `.locale` argument, and the locale would even be allowed to vary between columns. If no locale tag was supplied, `arrange()` would default to either `"en"` or `"C"` for the locale. This approach is relatively clean, but is practically very difficult because it would require cross package effort to generate and retain these locale tags. Additionally, it doesn't solve the problem of avoiding breakage for existing code that uses a non-English locale. Lastly, it would require an additional learning curve for users to understand how to use them in conjunction with `arrange()`.
