---
output: github_document
editor:
  markdown:
    wrap: sentence
    canonical: true
---

# Tidyup 7: Recoding and replacing values in the tidyverse

**Champion**: Davis

**Co-Champion**: Hadley

**Status**: Accepted

```{r}
library(dplyr, warn.conflicts = FALSE)
library(vctrs, warn.conflicts = FALSE)
```

## Abstract

Over the years, we have received numerous variations of the following two questions:

-   What is the best way to **recode** a column in the tidyverse?

-   What is the best way to **replace** a few values within a column in the tidyverse?

The point of this tidyup is to argue that there is a hole in dplyr's API surrounding these two questions, and to propose a solution that fills that hole by building on the intuition that people have around `case_when()`'s formula interface.

Let's first define exactly what we mean by recode and replace:

-   **Recoding** *creates a new vector* with an entirely new type.
    Unmatched locations that aren't handled by the recoding fall through to some default value, which uses a missing value if left unspecified.
    Output names are pulled from the names of the individual vectors that build the output values.

-   **Replacing** *modifies an existing vector*, retaining its type.
    Unmatched locations that aren't replaced retain the existing value of the original vector.
    Output names are left unchanged, regardless of whether the underlying value changed (mimicking `[<-` and `base::replace()`).

## Solution

In this tidyup, it's most compelling to outline the solution first so we can refer back to them when discussing motivation and showing examples.
The proposed solution is comprised of the following family of dplyr functions:

``` r
# - `...` are for two sided formulas
# - `...` are mutually exclusive with `from/to` (interactive vs programmatic API)

# Recoding

case_when(
  ...,
  .default = NULL,
  .unmatched = c("default", "error"),
  .ptype = NULL,
  .size = NULL
)

recode_values(
  x,
  ...,
  from = NULL,
  to = NULL,
  default = NULL,
  unmatched = c("default", "error"),
  ptype = NULL
)

# Replacing

replace_when(
  x,
  ...
)

replace_values(
  x,
  ...,
  from = NULL,
  to = NULL
)
```

`case_match()` will be superseded in favor of the more powerful and better named `recode_values()`.

For historical reasons, `case_when()` uses dotted argument names.
Ideally it would not, because the `...` of `case_when()` are not intended to be named, though we technically allow this right now.
We correct this historical mistake with the other functions in the family.
We anticipate this will make them easier to teach.

Powering this family are the following lower level vctrs functions, written in C for performance, and exposed in a package with minimal dependencies for use in other R packages where dplyr is too heavy of a dependency to take on.
These are not the focus of this tidyup, but it is worth knowing that they will also exist:

``` r
# - `...` must be empty, there is no formula API at this level

vec_case_when(
  cases,
  values,
  ...,
  default = NULL,
  unmatched = c("default", "error"),
  cases_arg = "cases",
  values_arg = "values",
  default_arg = "default",
  ptype = NULL,
  size = NULL,
  call = current_env()
)

vec_recode_values(
  x,
  ...,
  from,
  to,
  default = NULL,
  unmatched = c("default", "error"),
  from_as_list_of_vectors = FALSE,
  to_as_list_of_vectors = FALSE,
  x_arg = "x",
  from_arg = "from",
  to_arg = "to",
  default_arg = "default",
  ptype = NULL,
  call = current_env()
)

# Replacing

vec_replace_when(
  x,
  cases,
  values,
  ...,
  x_arg = "x",
  cases_arg = "cases",
  values_arg = "values",
  call = current_env()
)

vec_replace_values(
  x,
  ...,
  from,
  to,
  from_as_list_of_vectors = FALSE,
  to_as_list_of_vectors = FALSE,
  x_arg = "x",
  from_arg = "from",
  to_arg = "to",
  call = current_env()
)
```

It is straightforward to also construct `vec_if_else()` on top of `vec_case_when()` and expose this extremely useful utility within a low dependency package:

``` r
vec_if_else(
  condition,
  true,
  false,
  ...,
  missing = NULL,
  ptype = NULL,
  call = current_env()
)
```

## Implementation

Try for yourself at:

``` r
pak::pak("tidyverse/dplyr@feature/case-family")
```

Note that this implementation is not finished yet, and may have bugs.

-   [vctrs PR](https://github.com/r-lib/vctrs/pull/1984)

-   [dplyr branch](https://github.com/tidyverse/dplyr/tree/feature/case-family)

## Motivation

While dplyr's `case_when()` and `case_match()` are widely loved, over the years we have learned that there are still holes in the API that they are trying to fill:

-   They are designed to *create* new vectors, making them well suited for recoding, but poorly suited for replacing where you are *modifying* an existing vector.

-   They lack *programmatic* interfaces, like `plyr::mapvalues()` or `dplyr::recode(x, !!!cases)`.
    This is immediately noticeable when using a lookup table with `case_match()`.
    There is no intuitive way to define the lookup table at the top of a script and pass it through to the `case_match()` call, which we would encourage as a best practice if it was possible.

-   `case_match()` unfortunately has an unintuitive name.
    It tries to lean into the SQL "simple" `CASE WHEN` analogy combined with the fact that it works similarly to a `match()` call, but this doesn't land for most users, and many don't know it even exists.
    `recode_values()` and `replace_values()` are much more evocative names.

Below, we explore examples of both recoding and replacing (many extracted from real GitHub issues or discussions with users), comparing existing solutions to the proposed API.

### Recoding

#### As an inline lookup table with `case_match()`

Consider the following Likert scale scores.
We'd like to recode these from their numeric values to their character counterparts.

```{r}
data <- tibble(
  score = c(1, 2, 3, 4, 5, 2, 3, 1, 4)
)
```

To do this with existing tools, you'll likely be inclined to reach for `case_when()` or `case_match()`:

```{r}
data |>
  mutate(
    score = case_when(
      score == 1 ~ "Strongly disagree",
      score == 2 ~ "Disagree",
      score == 3 ~ "Neutral",
      score == 4 ~ "Agree",
      score == 5 ~ "Strongly agree"
    )
  )

# Or
data |>
  mutate(
    score = case_match(
      score,
      1 ~ "Strongly disagree",
      2 ~ "Disagree",
      3 ~ "Neutral",
      4 ~ "Agree",
      5 ~ "Strongly agree"
    )
  )
```

This is okay in some cases, but for a fixed Likert scale like this, it's typically more readable to extract the mapping out into a separate lookup table and refer to that within the `mutate()` call.
Unfortunately, `case_when()` and `case_match()` don't provide an easy way to do this, so you are forced to "inline" the lookup table.
With the newly proposed `recode_values()`, you can use `from` and `to` instead, which work similarly to `plyr::mapvalues()`:

```{r}
# Likert scale lookup table
# fmt: skip
likert_lookup <- tribble(
  ~from, ~to,
  1, "Strongly disagree",
  2, "Disagree",
  3, "Neutral",
  4, "Agree",
  5, "Strongly agree"
)

data |>
  mutate(
    score = recode_values(
      score,
      from = pull(likert_lookup, from),
      to = pull(likert_lookup, to)
    )
  )
```

It's quite difficult to achieve something similar with `case_match()`.
It requires a somewhat advanced level of formula manipulation for a fairly simple task:

```{r}
cases <- purrr::map2(likert_lookup$from, likert_lookup$to, \(from, to) {
  rlang::new_formula(from, to)
})

data |>
  mutate(score = case_match(score, !!!cases))
```

It's also common for a lookup table like `likert_lookup` to actually be defined in a separate CSV file, making it very awkward to inline it.
The `from/to` API makes this much easier:

```{r}
#| eval: false
likert_lookup <- read_csv("lookup.csv")

data |>
  mutate(
    score = recode_values(
      score,
      from = pull(likert_lookup, from),
      to = pull(likert_lookup, to)
    )
  )
```

Note that `recode_values()` still supports this "inline" lookup table via a formula API that is mutually exclusive with the `from` / `to` API.

```{r}
data |>
  mutate(
    score = recode_values(
      score,
      1 ~ "Strongly disagree",
      2 ~ "Disagree",
      3 ~ "Neutral",
      4 ~ "Agree",
      5 ~ "Strongly agree"
    )
  )
```

With the formula API, `recode_values()` is an exact replacement for `case_match()`, but comes with a powerful programmatic API as well.

#### As a lookup table with a join

More advanced users might skip `case_when()` and `case_match()` in favor of a join.
This isn't wrong, but it isn't an intuitive operation to reach for as a beginner, and typically requires extra steps if you want the resulting column to have the same name as the original one.

```{r}
# Likert scale
# fmt: skip
likert_lookup <- tribble(
  ~from, ~to,
  1, "Strongly disagree",
  2, "Disagree",
  3, "Neutral",
  4, "Agree",
  5, "Strongly agree"
)

# If you are okay with having both columns in the result, you stop here
left_join(data, likert_lookup, join_by(score == from))

# But often you want this
left_join(data, likert_lookup, join_by(score == from)) |>
  select(-score) |>
  rename(score = to)
```

With `recode_values()`, you recode directly within a `mutate()` call, which feels intuitive, and can assign directly back to the name of the column you recoded.

```{r}
data |>
  mutate(
    score = recode_values(
      score,
      from = pull(likert_lookup, from),
      to = pull(likert_lookup, to)
    )
  )
```

It's also worth mentioning that a join requires *data frames* as inputs, but recoding is really a *vector* level operation.
Note how `recode_values()` doesn't require data frames at all, making it suitable for package level programmatic usage as well.

```{r}
score <- data$score
from <- likert_lookup$from
to <- likert_lookup$to

recode_values(score, from = from, to = to)

# In this case you may consider using the vctrs variant as well, depending on
# your use case
vec_recode_values(score, from = from, to = to)
```

#### Accidentally dropping a value

When recoding a vector, it's entirely possible that you might accidentally miss a value.
This results in it being silently converted to a missing value.
Doubly confusing is that this missing value looks the same as preexisting missing values:

```{r}
data <- tibble(
  score = c(0, 1, 2, NA, 4, 5, 2, 3, 1, 4)
)

# Missed the `0`
data |>
  mutate(
    score_recoded = case_match(
      score,
      1 ~ "Strongly disagree",
      2 ~ "Disagree",
      3 ~ "Neutral",
      4 ~ "Agree",
      5 ~ "Strongly agree"
    )
  )
```

Some people guard against this by inserting and detecting a special `.default` value:

```{r}
#| error: true

data <- data |>
  mutate(
    score_recoded = case_match(
      score,
      1 ~ "Strongly disagree",
      2 ~ "Disagree",
      3 ~ "Neutral",
      4 ~ "Agree",
      5 ~ "Strongly agree",
      NA ~ NA,
      .default = "MISSED_ME"
    )
  )

if (any(data$score_recoded == "MISSED_ME")) {
  stop("Oh no!")
}
```

With the proposal in this tidyup, both `case_when()` and `recode_values()` would gain an `unmatched` argument which can optionally error on unmatched values rather than falling through to a `default`.
Note that you have to explicitly handle every value, even missing values, so they need to be included in your lookup table if you are using one.

```{r}
#| error: true

# Likert scale
# fmt: skip
likert_lookup <- tribble(
  ~from, ~to,
  1, "Strongly disagree",
  2, "Disagree",
  3, "Neutral",
  4, "Agree",
  5, "Strongly agree"
)

likert_lookup <- add_row(likert_lookup, from = NA, to = NA)

# Errors on the `0` that otherwise would fall through to `default`
data |>
  mutate(
    score = recode_values(
      score,
      from = pull(likert_lookup, from),
      to = pull(likert_lookup, to),
      unmatched = "error"
    )
  )
```

This is similar to using the join argument of the same name, but that requires a bit of mental gymnastics to get the right combination of `*_join()` and `unmatched` for this particular use case.
In this case, it's an `inner_join()` with `unmatched = c("error", "drop")` to declare that you're okay with unmatched keys from `y` dropping out, but you don't want any unmatched `x` keys.

```{r}
#| error: true

inner_join(
  data,
  likert_lookup,
  join_by(score == from),
  unmatched = c("error", "drop")
)
```

#### References

Questions and discussions related to recoding values which were referenced when creating this tidyup:

-   [Bsky conversation about this](https://bsky.app/profile/randvegan.bsky.social/post/3lsab7xfb6s2x){.uri}
-   [Libby's Linkedin post about this](https://www.linkedin.com/posts/libbyheeren_rstats-activity-7343291858275487744-XlPl?utm_source=share&utm_medium=member_desktop&rcm=ACoAAAy7IywB2qfaREGGoCca5XkthJ2hLjru6ts){.uri}
-   [Michael Chirico being confused about this](https://github.com/tidyverse/dplyr/issues/7694){.uri}
-   [A request for \`plyr::mapvalues()\`, which \`case_match()\` doesn't replicate](https://github.com/tidyverse/dplyr/issues/7027)
-   [Another request for \`plyr::mapvalues()\`, which \`case_match()\` doesn't replicate](https://github.com/tidyverse/dplyr/issues/5919#issuecomment-1943124726)
-   [Request for from/to style documentation examples pulled from recode()](https://github.com/tidyverse/dplyr/issues/6623)
-   [Me realizing that even \`vec_case_match()\` isn't perfect to replace \`mapvalues()\`](https://github.com/r-lib/vctrs/issues/1622)
-   [{funs} request for mapvalues()](https://github.com/tidyverse/funs/issues/15)
-   [{funs} issue about Lionel's vec_recode() attempt](https://github.com/tidyverse/funs/issues/29)
-   [coolbutuseless strict case_when](https://coolbutuseless.github.io/2018/09/06/strict-case_when/)

### Replacing

#### With `.default = col`

Traditionally, to replace a few values within a column in the tidyverse you'd use either `case_when()` or `case_match()` and set `.default = col` to retain the original vector's values in locations that you weren't replacing.

```{r}
# Collapse some, but not all, of these school names into common buckets
schools <- tibble(
  name = c(
    "UNC",
    "Chapel Hill",
    NA,
    "Duke",
    "Duke University",
    "UNC",
    "NC State",
    "ECU"
  )
)

schools |>
  mutate(
    name = case_match(
      name,
      c("UNC", "Chapel Hill") ~ "UNC Chapel Hill",
      c("Duke", "Duke University") ~ "Duke",
      .default = name
    )
  )
```

[This operation is so common](https://github.com/tidyverse/dplyr/issues/7696) that it feels like it deserves its own name that doesn't require you to set `.default = name`.
For that, this tidyup proposes `replace_values()`:

```{r}
schools |>
  mutate(
    name = replace_values(
      name,
      c("UNC", "Chapel Hill") ~ "UNC Chapel Hill",
      c("Duke", "Duke University") ~ "Duke"
    )
  )
```

Like `recode_values()`, `replace_values()` has an alternative `from` and `to` API that works well with lookup tables and allows you to move your replacement mapping up to the top of your script.

```{r}
# fmt: skip
schools_lookup <- tribble(
  ~from, ~to,
  "UNC", "UNC Chapel Hill",
  "Chapel Hill", "UNC Chapel Hill",
  "Duke", "Duke",
  "Duke University", "Duke"
)

schools |>
  mutate(
    name = replace_values(
      name,
      from = pull(schools_lookup, from),
      to = pull(schools_lookup, to)
    )
  )
```

An extremely neat feature of the `from` and `to` API is that they also take *lists* of vectors that describe the mapping, which has been designed to work elegantly with the fact that `tribble()` can create list columns, allowing you to further collapse this lookup table:

```{r}
# Condensed lookup table with a `many:1` mapping per row
# fmt: skip
schools_lookup <- tribble(
  ~from, ~to,
  c("UNC", "Chapel Hill"), "UNC Chapel Hill",
  c("Duke", "Duke University"), "Duke"
)

# Note the type of `from` here
schools_lookup
schools_lookup$from

# Works the same as before
schools |>
  mutate(
    name = replace_values(
      name,
      from = pull(schools_lookup, from),
      to = pull(schools_lookup, to)
    )
  )
```

#### With the native pipe

Using `case_match(x, .default = x)` has [an annoying issue](https://github.com/tidyverse/dplyr/issues/6962) when the native pipe is involved:

```{r}
#| error: true

# Let's assume we have a little preprocessing to do before we can use
# `case_match()`. We need to strip off the `Program-` prefix for some
# of the schools.

schools <- tibble(
  name = c(
    "Program-UNC",
    "Chapel Hill",
    NA,
    "Duke",
    "Duke University",
    "UNC",
    "Program-NC State",
    "ECU"
  )
)

# Pipe placeholder can only appear once, so this is invalid R code
schools |>
  mutate(
    name = name |>
      stringr::str_remove("Program-") |>
      case_match(
        .x = _,
        c("UNC", "Chapel Hill") ~ "UNC Chapel Hill",
        c("Duke", "Duke University") ~ "Duke",
        .default = _
      )
  )
```

Because the pipe placeholder of `_` can only appear once per function call, this is invalid R code!
`replace_values()` happens to solve this problem elegantly by only requiring you to specify the column one time.
In fact, because the pipe placeholder is automatically assigned to the first input, you can remove the `_` altogether.

```{r}
schools |>
  mutate(
    name = name |>
      stringr::str_remove("Program-") |>
      replace_values(
        c("UNC", "Chapel Hill") ~ "UNC Chapel Hill",
        c("Duke", "Duke University") ~ "Duke"
      )
  )
```

#### Type stability

One subtle issue with using `case_when(.default = x)` or `case_match(.default = x)` is that it isn't type stable on `x`, which you probably do want when you're just replacing a few values.

```{r}
# Pretend it is important for this to be an <integer> column
x <- c(1L, 2L, 3L)

# Replacing `1` with `0` and retaining `x` as the default results in a <double>
# vector, because the common type of `x` (integer) and `0` (double) is double.
# We've lost the original <integer> type.
typeof(case_match(x, 1 ~ 0, .default = x))
```

Remember that `case_when()` and `case_match()` were originally designed to *create totally new vectors*, so they aren't meant to be type stable on their inputs like you might expect here.
You can force type stability, but it's a bit of work:

```{r}
typeof(case_match(x, 1 ~ 0, .default = x, .ptype = x))
```

`replace_values()` is an all-around better solution here, it's type stable by default, and much shorter!

```{r}
typeof(replace_values(x, 1 ~ 0))
```

This is particularly useful when partially recoding factors to existing levels.

```{r}
pets <- tibble(
  name = c("Max", "Bella", "Chuck", "Luna", "Cooper"),
  # Note the so-far unused "puppy" level:
  type = factor(
    c("dog", "dog", "cat", "dog", "cat"),
    levels = c("dog", "cat", "puppy")
  ),
  age = c(1, 3, 5, 2, 4)
)

# Recode some values to `"puppy"` using a character on the RHS
pets |>
  mutate(
    type = type |>
      replace_when(type == "dog" & age <= 2 ~ "puppy")
  )
```

```{r}
#| error: true

# Note the type safety! Only existing levels can be used here.
pets |>
  mutate(
    type = type |>
      replace_when(type == "dog" & age <= 2 ~ "pup")
  )
```

The corresponding call to `case_when()` is much less intuitive due to the fact that the common type of character and factor is character, which means that the output of this `case_when()` call would be a character without the `.ptype` argument to force it to be a factor.

```{r}
pets |>
  mutate(
    type = case_when(
      type == "dog" & age <= 2 ~ "puppy",
      .default = type,
      .ptype = type
    )
  )
```

#### Retaining names

Another subtle difference between *recoding* and *replacing* is where the output names are pulled from.
With recoding, you create an entirely new vector, so output names are built from the input vectors that end up building the result:

```{r}
x <- c(a = 1, b = 2, c = 3)

from <- c(2, 3)
to <- c(x = 20, y = 30)

recode_values(x, from = from, to = to)
```

With replacing, you modify an *existing* vector, so output names are the exact same as the original vector, regardless of whether you modified the underlying value or not.

```{r}
replace_values(x, from = from, to = to)
```

This is the exact principle used by `base::replace()` and `[<-`:

```{r}
# Notice how `b` stays even though `20` takes the place of `2`
replace(x, 2, c(x = 20))
```

#### Retaining intent

When you're replacing a few values in one column based on a condition computed from another, you typically reach for `case_when()`.
However, because you provide the column as the last argument (the `.default`), the intent of the operation is lost.
With `replace_when()`, you specify the column as the first argument, which helps you retain a natural reading order and matches a common tidyverse pattern where the first input usually correlates strongly to the type, size, and values of the output.
It also works nicely with the pipe!

```{r}
# Svalbard and Jan Mayen technically have the same ISO code, but for our
# analysis assume we need them to be different
data <- tibble(
  country = c("USA", "Svalbard", "Jan Mayen", "Canada"),
  iso_code = c("US", "SJ", "SJ", "CA")
)

# The intent to just replace a few values in `iso_code` gets lost
# due to the fact that it is specified as the last argument
data |>
  mutate(
    iso_code = case_when(
      country == "Svalbard" ~ "SJ-Svalbard",
      country == "Jan Mayen" ~ "SJ-Jan Mayen",
      .default = iso_code
    )
  )

# `replace_when()` allows you to pull it to the front, or even pipe it in!
# This makes the intent of your code much clearer.
data |>
  mutate(
    iso_code = iso_code |>
      replace_when(
        country == "Svalbard" ~ "SJ-Svalbard",
        country == "Jan Mayen" ~ "SJ-Jan Mayen"
      )
  )
```

#### `na_if()` alternative

A fun result that falls out of `replace_values()` is that it's a more flexible (and likely more intuitive) alternative to `na_if()`:

```{r}
x <- c(1, 2, 0, -99, 12)

# To convert `0` and `-99` to `NA`, you have to do it in two calls
x |> na_if(0) |> na_if(-99)

x |> replace_values(from = c(0, -99), to = NA)
```

#### `coalesce()` alternative

It's also interesting to think of `replace_values()` as a simple `coalesce()` alternative:

```{r}
x <- c(1, 2, NA, 3, NA, 5)

coalesce(x, 0)

replace_values(x, NA ~ 0)
```

#### `tidyr::replace_na()` alternative

`replace_values()` is also a generalization of `tidyr::replace_na()`:

```{r}
data <- tibble(
  x = c(1, 2, NA, 4, NA),
  y = c(NA, "a", NA, "b", NA)
)

tidyr::replace_na(data, list(x = 0, y = "unknown"))

data |>
  mutate(
    x = tidyr::replace_na(x, 0),
    y = tidyr::replace_na(y, "unknown")
  )

data |>
  mutate(
    x = replace_values(x, NA ~ 0),
    y = replace_values(y, NA ~ "unknown")
  )
```

A similar comparison can be done for `naniar::replace_na_with()` and `naniar::replace_with_na()`.

#### Multiple columns

In some cases, you may need to replace values in multiple columns at once.
In our experience, this isn't quite as common as the other examples, but it's still possible using the fact that `replace_values()` and friends can take data frames as inputs, and that `mutate()` will automatically unpack unnamed data frames:

```{r}
data <- tibble(
  x = c(2, 5, 12, 15, 18, 3, 3, 7),
  y = c(10, 5, 7, 1, 6, 4, 3, 9),
  age = c(20, 15, 25, 45, 38, 41, 45, 46)
)

data |>
  mutate(replace_when(
    pick(x, y),
    age < 20 ~ tibble(x = 0, y = 0),
    age > 40 ~ tibble(x = NA, y = NA)
  ))
```

#### References

Questions and discussions related to replacing values which were referenced when creating this tidyup:

-   [2018 - Conditionally mutate selected rows, our 2nd oldest open issue](https://github.com/tidyverse/dplyr/issues/4050){.uri}
-   [{funs} attempt at recode_when()](https://github.com/tidyverse/funs/pull/66)
-   [Updating multiple values at once in \`na_if()\`](https://github.com/tidyverse/dplyr/issues/7651)
-   [A request for "case_replace()"](https://github.com/tidyverse/dplyr/issues/7696)
-   [\`case_match(.x, .default = .x)\` doesn't play nicely with pipes! Underrated problem IMO.](https://github.com/tidyverse/dplyr/issues/6962)
-   [mutate_where() twitter request](https://x.com/antoine_fabri/status/1392127389195452416){.uri}

## Backwards compatibility

### `case_match()`

`case_match()` will be superseded in favor of `recode_values()`, a more powerful and better named alternative.
All existing code using `case_match()` will continue to work.

## How to teach

When teaching these new functions, it's probably best to introduce them in pairs, i.e. `case_when() / replace_when()` and `recode_values() / replace_values()`, but within the same overarching lesson.

One key thing to avoid is talking about `dplyr::recode()`.
It is unfortunate that this function shares such a similar name as `recode_values()`, but `dplyr::recode()` has been superseded for a while now, and the API of `recode_values()` is much better.
In particular, the pattern of splicing in a named vector to `dplyr::recode()` using `!!!` is now much more elegantly handled using `from` and `to` arguments of `recode_values()`.

```{r}
set.seed(1234)

x <- sample(c("a", "b", "c"), 6, replace = TRUE)

lookup <- c(
  a = "Apple",
  b = "Banana",
  c = "Candy"
)

dplyr::recode(x, !!!lookup)

lookup <- tibble::enframe(lookup, "from", "to")

dplyr::recode_values(x, from = lookup$from, to = lookup$to)
```

## Additional considerations

### dbplyr

`case_when()` already nicely translates to SQL's "searched" `CASE WHEN`, and `case_match()` already translates to SQL's "simple" `CASE WHEN`.
`recode_values()` supersedes `case_match()` with a similar API, so we anticipate that it should also translate elegantly to a "simple" `CASE WHEN` statement.

Both `replace_when()` and `replace_values()` could be treated as `case_when()` and `recode_values()` calls where you already know what the fallthrough `ELSE` condition is - the original column.
This wouldn't be type stable on the original column type though, so we may also want to look into translating to `UPDATE + SET + WHERE` alongside the simple `CASE WHEN`, like [this example](https://stackoverflow.com/questions/15766102/i-want-to-use-case-statement-to-update-some-records-in-sql-server-2005).

The `from` and `to` API of `recode_values()` and `replace_values()` could likely be programmatically translated to their equivalent formula API, and run back through the existing code that handles the formula API for these functions.

It is unlikely that the new `unmatched` argument could be supported.

## Appendix

### References

Sources of inspiration considered while designing these APIs:

-   [naniar](https://naniar.njtierney.com/)
-   [tidyr::replace_na()](https://tidyr.tidyverse.org/reference/replace_na.html)
-   [base::replace()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/replace.html)
-   [plyr::mapvalues()](https://cran.r-project.org/web/packages/plyr/refman/plyr.html#mapvalues)
-   [kit::vswitch()](https://cran.r-project.org/web/packages/kit/refman/kit.html#vswitch+2Fnswitch)
-   [Excel's XLOOKUP](https://support.microsoft.com/en-us/office/xlookup-function-b7fd680e-6d10-43e6-84f9-88eae8bf5929)
-   [Stata's recode](https://www.stata.com/manuals/drecode.pdf)

### Full API

While working on this tidyup, we developed the following table of possible ways you might want to use these functions to ensure that we captured every possibility we could think of.

+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Usage        | Action  | Style   | RHS    | Function                                                       | Notes                                  |
+==============+=========+=========+========+================================================================+========================================+
| Interactive  | Create  | if/else | 1      | case_when(condition \~ "value", condition2 \~ "value2")        | Exists                                 |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Interactive  | Create  | if/else | N      | case_when(condition \~ column, condition2 \~ column2)          | Exists                                 |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Interactive  | Create  | switch  | 1      | recode_values(x, 1 \~ "value1", 2 \~ "value2")                 | case_match()                           |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Interactive  | Create  | switch  | N      | recode_values(x, 1 \~ column, 2 \~ column2)                    | case_match()                           |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Interactive  | Replace | if/else | 1      | replace_when(x, condition \~ 0, condition2 \~ NA)              | Conditional mutate()                   |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Interactive  | Replace | if/else | N      | replace_when(x, condition \~ column, condition2 \~ column2)    | Conditional mutate()                   |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Interactive  | Replace | switch  | 1      | replace_values(x, 1 \~ 0, 2 \~ NA)                             | Conditional mutate()                   |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Interactive  | Replace | switch  | N      | replace_values(x, 1 \~ column, 2 \~ column2)                   | Conditional mutate()                   |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Programmatic | Create  | if/else | 1      | vec_case_when(list(case, case2), list("value", "value2"))      | numpy.select()                         |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Programmatic | Create  | if/else | N      | vec_case_when(list(case, case2), list(column, column2))        | numpy.select()                         |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Programmatic | Create  | switch  | 1      | recode_values(x, from = c(1, 2), to = list("value", "value2")) | plyr::mapvalues() + type change        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | Convenience API:                                               |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | recode_values(x, from = c(1, 2), to = c("value", "value2"))    |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | vctrs API:                                                     |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | vec_recode_values(x, from, to)                                 |                                        |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Programmatic | Create  | switch  | N      | recode_values(x, from = c(1, 2), to = list(column, column2))   |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | vctrs API:                                                     |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | vec_recode_values(x, from, to, to_as_list_of_vectors = TRUE)   |                                        |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Programmatic | Replace | if/else | 1      | vec_replace_when(list(case, case2), list(0, NA))               |                                        |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Programmatic | Replace | if/else | N      | vec_replace_when(list(case, case2), list(column, column2))     |                                        |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Programmatic | Replace | switch  | 1      | replace_values(x, from = c(1, 2), to = list(0, NA))            | plyr::mapvalues(), na_if() alternative |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | Convenience API:                                               |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | replace_values(x, from = c(1, 2), to = c(0, NA))               |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | vctrs API:                                                     |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | vec_replace_values(x, from, to)                                |                                        |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
| Programmatic | Replace | switch  | N      | replace_values(x, from = c(1, 2), to = list(column, column2))  |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | vctrs API:                                                     |                                        |
|              |         |         |        |                                                                |                                        |
|              |         |         |        | vec_replace_values(x, from, to, to_as_list_of_vectors = TRUE)  |                                        |
+--------------+---------+---------+--------+----------------------------------------------------------------+----------------------------------------+
